Automatically Splitting a Two-Stage Lambda Calculus

### 2015-07-10 paper/examples.tex
Remove graphics from intro. Move to examples.

A major motivation for staged programming in computer graphics is to express
logic related to a single application concern conveniently as a single staged
function (referred to as a pipeline program)\,\cite{Foley:2011}, rather than
as a collection of functions distributed across different GPU pipeline stages.

Pipeline programs can be directly expressed as staged functions in \lang\, with splitting 
used to generate monostaged code required for each GPU pipeline stage
(as well as the plumbing to communicate values between the stages).
Composition of such functions in \lang\ allows more complex pipelines to be expressed in a modular programming style
that is becoming increasingly desirable in graphics shading languages\,\cite{Foley:2011,He:2014}. A simple example of composition of staged functions (representing GPU pipeline programs) is given below.  The code defines a pipeline as a staged function that takes an object definition in the first stage (\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second stage and emits the color of the object at the specified pixel. The code defines a pipeline combinator \texttt{shade} which modulates the results of the object's reflectance at a given pixel with the results of a function that computes the object's albedo (surface color).  
%
\begin{lstlisting} 
1`datatype object = ...`
2`atsignnext{
  type coord = int * int
  type color = int
}`
1`type pipeline = object * $`2`coord`1` -> `$2`color`

1`fun shade (refl : pipeline,
           albedo : pipeline) : pipeline =
  (fn (obj : object, next{2`xy`} : $2`coord`) =>
       2`prev{`1`refl   (obj,next{2`xy`})`2`}
     * prev{`1`albedo (obj,next{2`xy`})`2`}`)
\end{lstlisting}
%
Most pipeline programming models either force the user to manually split their programs (which harms modularity).
Others, such as \cite{Foley:2011}, offer only a few multistage types (namely, first order functions) and give 
combinators like \texttt{par} as language features.
In contrast, \lang\ is powerful enough to define those combinators in-language.

>>>>

Programs featuring staged structure are particularly important in real-time
computer graphics since modern graphics architectures \emph{require} that
graphics computations be organized into distinct {\em passes} which perform
increasingly fine-grained computations (e.g., per object, per screen region,
per pixel) where many computations in later passes use the
results from a single computation in an earlier pass\,\cite{OpenGL4Spec}.

While performant, programming with explicit passes results in complex
code where invariants must hold across different passes, and local
changes for one pass may require changes to other passes to ``plumb''
its inputs through all upstream passes. Thus, requiring manual
separation of passes breaks key programming abstractions such as
composition and modularity.  Graphics researchers therefore have
suggested using explicitly staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014} to express logic related to
a single application concern conveniently as a single staged function
(referred to as a pipeline program)\,\cite{Foley:2011}.

Pipeline programs can be directly expressed as staged functions in
\lang\ with splitting used to generate monostaged code required for
each GPU pipeline stage. Composition of such functions in
\lang\ allows more complex pipelines to be expressed in a modular
programming style that is becoming increasingly desirable as graphics
programs increase in complexity\,\cite{Foley:2011,He:2014}.

A simple example of composition of staged functions (representing GPU
pipeline programs) is given below.  The code defines a pipeline as a
staged function that takes an object definition in the first stage
(\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second
stage and emits the color of the object at the specified pixel. The
code defines a pipeline combinator \texttt{shade} which modulates the
results of the object's reflectance at a given pixel with the results
of a function that computes the object's albedo (surface color).
%
\begin{lstlisting} 
1`datatype object = ...`
2`atsignnext{
  type coord = int * int
  type color = int
}`
1`type pipeline = object * $`2`coord`1` -> `$2`color`

1`fun shade (refl : pipeline,
           albedo : pipeline) : pipeline =
  (fn (obj : object, next{2`xy`} : $2`coord`) =>
       2`prev{`1`refl   (obj,next{2`xy`})`2`}
     * prev{`1`albedo (obj,next{2`xy`})`2`}`)
\end{lstlisting}
%
Whereas prior systems provides combinators like \texttt{par} as
language features\,\cite{Foley:2011} to support composition, \lang\ is
powerful enough to define those combinators in-language.

### 2015-07-10 paper/intro.tex
Remove graphics from intro. Move to examples.

Consider a function $F$ which takes an $x$ and $y_1,\dots,y_m$, and computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_1, y_2 \ldots, y_m)$@ = 
  @$f(x, y_1);  f(x, y_2); \ldots; f(x, y_m)$@
\end{lstlisting}
%
In this common scenario, a multivariate function is applied many times with one
argument fixed, so it is advantageous to find the computations in $f$ that
depend on $x$ and perform them only once. This is the intent of compiler
optimization techniques like loop hoisting and common subexpression elimination,
which identify the computations common to all invocations of $f$, and perform
them once at the beginning of the computation.

Programs featuring this structure are particularly important in real-time
computer graphics; in fact, modern graphics architectures \emph{require} that
graphics computations be organized into distinct {\em passes} which perform
increasingly fine-grained computations (for example, per object, per region of
the screen, and per pixel, where many computations in later passes use the
results from a single computation in an earlier pass)\,\cite{OpenGL4Spec}.
While performant, this paradigm of programming with explicit passes results in
complex code where invariants must hold across different passes, and local
changes for one pass may require changes to other passes to ``plumb'' inputs
from all other frequencies (\TODO explain what ``frequency'' means here)\,\cite{Foley:2011}.
%
In other words, requiring manual separation of passes breaks key programming
abstractions such as composition and modularity.
%
Graphics researchers therefore have suggested using explicitly staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
mechanical task of separating such programs into passes.

Programming languages researchers have studied similar problems. In the 1980s,
J{\o}rring and Scherlis identified {\em frequency reduction} and {\em
precomputation}~\cite{JS86-staging} as mechanisms for efficiently evaluating
staged computations. To perform frequency reduction, one identifies and hoists
computations that are performed multiple times, in order to compute them only
once. To perform precomputation, one identifies computations that can be
performed in advance, and does so---for example, at compile time if the relevant
inputs are statically known.

There has been much work on language-based techniques for performing
precomputation. Partial evaluation~\cite{futamura71,jones96} achieves
precomputation by specializing functions to specific arguments. Going back to
our example, if $f$ specializes to a particular $v$, written $f_v$, such that
$f(v,y) = f_v(y)$, then $F$ can be specialized to $v$ as
\begin{lstlisting}
fun @$F_v(y_1, y_2, \ldots, y_m)$@ = 
  @$f_v(y_1);  f_v(y_2); \ldots; f_v(y_m)$@.
\end{lstlisting}
eliminating the need to compute $m$ times those parts of $f(v,-)$ which do not
depend on the second argument.

A closely related technique, {\em metaprogramming}, allows writing
programs that generate specialized
code~\cite{davies96,Taha97,DP01-modal,NP05-nn}. Metaprogramming
enables fine-grained control over specialization by requiring \emph{staging
annotations} that explicitly mark the stage of each computation. (In
partial evaluation, we typically assume that a \emph{binding-time
  analysis} has automatically determined which parts of $f$ can be
specialized.)

While partial evaluation and metaprogramming have been studied extensively,
frequency reduction techniques have not seen the same attention. In
\cite{JS86-staging}, J{\o}rring and Scherlis proposed {\em pass separation} as a
manual method for {\em splitting} a program into multiple passes.
%
In our example, pass separation would split the function $f$ into two passes
$f_1$ and $f_2$ such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating the first pass on $x$, and using the result $z$ to evaluate the
second pass on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{multipass}(x, y_1, y_2, \ldots, y_m)$@ = 
  let @$z = f_1(x)$@
  in @$f_2(z, y_1);  f_2(z, y_2); \ldots; f_2(z, y_m)$@
\end{lstlisting}
%
The key difference between pass separation and partial evaluation (or
metaprogramming) is that the former can be performed without access to
the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
while $F_v$ is defined only for $x=v$.  Note that unlike
precomputation, frequency reduction does not require knowledge of
arguments; however, if a particular $v$ is known, we can use partial evaluation
to specialize $F_\textrm{multipass}$ to that $v$.

Prior work on partial evaluation and metaprogramming has shown that
precomputation can largely be automated for functional languages. In contrast,
automatic pass separation has been limited to simple languages that do not
support recursion or first-class
functions~\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}. In this paper, we
extend this work by presenting a pass separation algorithm for a typed lambda
calculus.

As our starting point, we consider an explicitly-two-staged, typed lambda
calculus \lang\ in the style of Davies~\cite{davies96}, with a $\fut$ modality
denoting computation in a later stage (\ref{sec:semantics}). We then present a
splitting algorithm to pass-separate a \lang\ program into two passes
corresponding to the stages of computation in that program
(\ref{sec:splitting}). These passes are monostaged programs, expressed in a
conventional functional language.

The most interesting case involves splitting a mixed-stage function (like $f$ in
the example) which is moreover recursive. This yields a first pass function
$f_1$ which computes from its first-stage argument $x$ a \emph{boundary} data
structure encoding the interstage communication in $f(x,-)$. When $f$ recurs on
$x$, this boundary structure itself has a recursive type.
%
The second pass is a function $f_2$ which takes the boundary data structure
$f_1(x)$ and a second-stage argument $y_i$, and completes the computation
$f(x,y_i)$, traversing the boundary structure in light of the particular
argument $y_i$.

@@

%The crux of the splitting algorithm is splitting recursive mixed stage
%functions that involve both a first stage and second stage computations.  When
%split, such a function turns into a first-pass function that yields a {\em
%boundary value} that encodes the ``first-stage'' part of the computation.  If
%the function recurs on a first-stage argument, then the boundary value is
%typically a recursive data structure.  In other words, the splitting algorithm
%maps computational recursion into a recursive data type. If the boundary
%structure is generated as a result of a conditional, it is typically encoded as
%a (recursive) sum type, such as a tree, where the kind of the node indicates
%the recursion status of the first-stage evaluation.  The second-pass part of
%the computation, which corresponds to the ``second stage'' takes the boundary
%value as an argument as well as the second-stage argument and uses them to
%complete the computation, often by traversing the now precomputed boundary in
%light of the now available second-stage argument.

>>>>

Consider a function $F$ which takes an $x$ and $y_1,\dots,y_m$, and computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_1, y_2 \ldots, y_m)$@ = 
  @$f(x, y_1);  f(x, y_2); \ldots; f(x, y_m)$@
\end{lstlisting}
%
In this scenario, a multivariate function is applied many times with
one argument fixed, so it is advantageous to find the computations in
$f$ that depend on $x$ and then stage program execution so they are
performed only once at the beginning of execution.

%Programs featuring this structure are particularly important in real-time
%computer graphics; in fact, modern graphics architectures \emph{require} that
%graphics computations be organized into distinct {\em passes} which perform
%increasingly fine-grained computations (for example, per object, per region of
%the screen, and per pixel, where many computations in later passes use the
%results from a single computation in an earlier pass)\,\cite{OpenGL4Spec}.
%While performant, this paradigm of programming with explicit passes results in
%complex code where invariants must hold across different passes, and local
%changes for one pass may require changes to other passes to ``plumb'' inputs
%from all other frequencies (\TODO explain what ``frequency'' means here)\,\cite{Foley:2011}.
%In other words, requiring manual separation of passes breaks key programming
%abstractions such as composition and modularity.
%Graphics researchers therefore have suggested using explicitly staged programs
%\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
%mechanical task of separating such programs into passes.

%Programming languages researchers have studied similar problems, with
%J{\o}rring and Scherlis identifying {\em frequency reduction} and {\em
% precomputation} as mechanisms for efficiently evaluating staged
%computations\,\cite{JS86-staging}.

J{\o}rring and Scherlis classify automatic program staging
transformations, such as the one described above, as forms of {\em
  frequency reduction} or {\em
  precomputation}\,\cite{JS86-staging}. To perform frequency
reduction, one identifies and hoists computations that are performed
multiple times, in order to compute them only once. To perform
precomputation, one identifies computations that can be performed in
advance, and does so---for example, at compile time if the relevant
inputs are statically known.

One common precomputation technique is partial
evaluation\,\cite{futamura71,jones96}, which relies on dynamic
compilation to specialize functions to known argument values. Going
back to our example, if $f$ specializes to a particular $v$, written
$f_v$, such that $f(v,y) = f_v(y)$, then $F$ can be specialized to $v$
as
\begin{lstlisting}
fun @$F_v(y_1, y_2, \ldots, y_m)$@ = 
  @$f_v(y_1);  f_v(y_2); \ldots; f_v(y_m)$@.
\end{lstlisting}
eliminating the need to compute $m$ times those parts of $f(v,-)$
which do not depend on the second argument.

Closely related to partial evaluation is {\em metaprogramming}, where
known values represent program code to be executed in a later stage
\,\cite{davies96,Taha97,DP01-modal,NP05-nn}. Metaprogramming enables
fine-grained control over specialization by requiring explicit
\emph{staging annotations} that mark the stage of each expression.

%(In partial evaluation, we typically assume that a
%\emph{binding-time analysis} has automatically determined which parts
%of $f$ can be specialized.)

%This is the intent of compiler optimization techniques like loop
%hoisting and common subexpression elimination, which identify the
%computations common to all invocations of $f$, and perform them once
%at the beginning of the computation.

%While partial evaluation and metaprogramming have been studied
%extensively, frequency reduction techniques have not seen the same
%attention.

While simple forms of frequency reduction include standard compiler
optimizations such as loop hoisting and common subexpression
elimination, J{\o}rring and Scherlis proposed the more general
transformation of {\em splitting} a program into multiple passes
(called pass separation in \,\cite{JS86-staging}). In our example,
splitting transforms the function $f$ into two passes $f_1$ and $f_2$
such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating the first pass on $x$, and using the result $z$ to evaluate
the second pass on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{multipass}(x, y_1, y_2, \ldots, y_m)$@ = 
  let @$z = f_1(x)$@
  in @$f_2(z, y_1);  f_2(z, y_2); \ldots; f_2(z, y_m)$@
\end{lstlisting}
%
The key difference between splitting and partial evaluation (or
metaprogramming) is that the former can be performed without access to
the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
while $F_v$ is defined only for $x=v$.  Therefore, unlike partial
evaluation, splitting is a static program transformation
(``metastatic'' in parial evaluation terms) and does not require
dynamic code generation.

% Note that unlike precomputation, frequency reduction does not require
%knowledge of arguments; however, if a particular $v$ is known, we can
%use partial evaluation to specialize $F_\textrm{multipass}$ to that
%$v$.

Prior work on partial evaluation and metaprogramming has demonstrated
automatic application of these techniques on higher order functional
languages. In contrast, automatic splitting transforms have been
limited to simple languages that do not support recursion or
first-class
functions\,\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}. In
this paper, we extend this work by presenting a splitting algorithm
for a typed lambda calculus.

As our starting point, we consider an explicitly-two-staged, typed
lambda calculus \lang\ in the style of Davies\,\cite{davies96}, with a
$\fut$ modality denoting computation in a later stage
(\ref{sec:semantics}). We then present a splitting algorithm that
transforms \lang\ program into two passes corresponding to the stages
of computation in that program (\ref{sec:splitting}). These passes are
monostaged programs, expressed in a conventional functional language.

The most interesting case involves splitting a recursive mixed-stage
function (like $f$ in the example). This yields a first pass function
$f_1$ which computes from its first-stage argument $x$ a
\emph{boundary} data structure encoding the interstage communication
in $f(x,-)$. When $f$ recurs on $x$, this boundary structure itself
has a recursive type.  The second pass is a function $f_2$ which takes
the boundary data structure $f_1(x)$ and a second-stage argument
$y_i$, and completes the computation $f(x,y_i)$, traversing the
boundary structure in light of the particular argument $y_i$.

@@

### 2015-07-10 paper/relatedwork.tex
Remove graphics from intro. Move to examples.

\emph{Program bifurcation}\,\cite{DeNiel91} (also called \emph{binding time separation}\,\cite{Mogensen89a})
is a metastatic staging transformations
used as a preprocessing step to simplify the specialization phase of self-applied partial evaluators. 
Bifurcation is similar to stage splitting in that it factors the input program
into two versions based on the stages of the input.
However, in bifurcation there is no building of data structures to be use in the resumer;
instead, the second stage version of the code also has 
access to the first stage inputs and just recomputes whatever it needs.
That difference aside, the grammar-based binding-time specifications used in bifurcation
were capable of descring data structures with purely-static, purely-dynamic, and mixed stage content,
much like the type system of \lang.

@@

Splitting algorithms have also been a topic of interest in
programming systems for computer graphics, where, to achieve
high performance, programs are typically manually separated into
components by frequency of execution (Indeed, separation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  The software engineering
challenges of modifying multiple per-stage programs have led to
suggestions of writing graphics programs in an explicitly-staged
programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
deferring the task of pass separation to the compiler. However, all
prior efforts, like that of Knoblock and Ruf\,\cite{knoblock96} have
been limited to simple, imperative languages.

>>>>

\emph{Program bifurcation}\,\cite{DeNiel91} (also called \emph{binding time separation}\,\cite{Mogensen89a})
is a metastatic staging transformations
used as a preprocessing step to simplify the specialization phase of self-applied partial evaluators. 
Bifurcation is similar to stage splitting in that it factors the input program
into two versions based on the stages of the input.
However, in bifurcation there is no building of data structures to be use in the resumer;
instead, the second stage version of the code also has 
access to the first stage inputs and just recomputes whatever it needs.
That difference aside, the grammar-based binding-time specifications used in bifurcation
were capable of describing data structures with purely-static, purely-dynamic, and mixed stage content,
much like the type system of \lang.

@@

Splitting algorithms have also been a topic of interest in programming
systems for computer graphics, where, to achieve high performance,
programs are typically manually separated into components by frequency
of execution corresponding to support graphics hardware pipeline
stages. The software engineering challenges of modifying multiple
per-stage programs have led to suggestions of writing graphics
programs in an explicitly-staged programming
language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and deferring the
task of pass separation to the compiler. However, all prior efforts,
like that of Knoblock and Ruf\,\cite{knoblock96} have been limited to
simple, imperative languages.
